---
title: "Running meta-analyses with miniMeta"
author: "Theodore Lytras \\<thlytras@gmail.com\\>"
date: "`r Sys.Date()`"
version: 0.0.1
output: pdf_document
vignette: >
  %\VignetteIndexEntry{miniMeta_manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

[miniMeta](https://github.com/thlytras/miniMeta) 
is a [shiny](https://shiny.rstudio.com/) web application that makes 
it easy to run meta-analyses, without having any familiarity with R 
and its console. 
It is intended for use by non-statistical colleagues, or as educational tool.

miniMeta can be run locally from an R installation, or fully online 
from a shiny server. It provides a "live" web interface where the 
meta-analysis is automatically updated as data are entered or options are adjusted. 
It can save (export) the results into files (RDS files) which can be
later loaded (imported) to miniMeta, or also further processed in R if more
detailed analyses are required.

Under the hood miniMeta relies on the 
[meta](https://CRAN.R-project.org/package=meta) package for R, and can be
thought of as a graphical front-end for it. 
Analysis and plotting is done by functions from meta, in particular 
[`metabin()`](https://www.rdocumentation.org/packages/meta/versions/4.9-7/topics/metabin), 
[`metagen()`](https://www.rdocumentation.org/packages/meta/versions/4.9-7/topics/metagen) and 
[`forest.meta()`](https://www.rdocumentation.org/packages/meta/versions/4.9-7/topics/forest.meta); 
see the associated function documentation for details.


# How to install and run locally

Open R, install package [devtools](https://github.com/thlytras/devtools) if 
you don't already have it, and then use `install_github()` to install the 
[miniMeta](https://github.com/thlytras/miniMeta) package from GitHub:

```{r, eval=FALSE, echo=TRUE}
install.packages("devtools")
devtools::install_github("thlytras/miniMeta")
```

This will bring in from CRAN all required dependencies if not already 
installed, such as packages 
[shiny](https://CRAN.R-project.org/package=shiny), 
[meta](https://CRAN.R-project.org/package=meta) and 
[readxl](https://CRAN.R-project.org/package=readxl).

In order to run miniMeta, load the package with `library()` and launch it
with `miniMeta()`:

```{r, eval=FALSE, echo=TRUE}
library(miniMeta)
miniMeta()
```

This will run miniMeta in a new browser window.


# Using miniMeta

At the top of the miniMeta window, there are three tabs named "RCT module",
"Observational studies module" and "Tools" respectively. 
Click one of the first two if you want to run a meta-analysis of 
Randomized Controlled Trials (RCTs) or observational studies. 
In the "Tools" tab there is a suite of small tools
(currently just one, for making adjusted indirect comparisons) 
with which to perform various associated calculations.


## Randomized Controlled Trials (RCT) module

At the top, there are buttons "Import meta-analysis" and "Export meta-analysis".
Click "Export meta-analysis" to save the current meta-analysis (including 
dataset, analysis options and plot options) as an RDS file. 
This can be loaded later into miniMeta by clicking "Import meta-analysis". 

The data are input on the left section of the page, and results are output on
the right section. Options for the analysis are also set on the left section, 
whereas options for the forest plot are set on the right section.

### Input section (left side)

Data can be input in two ways: by hand at the provided table widget, or loaded 
from an Excel spreadsheet. The Excel file must have the exact same layout 
as the table widget: each row represents one study, the first column holds 
the study labels, followed by the number of events (number of participants 
with the outcome) and total number of participants randomized in the 
intervention arm, the number of events and participants randomized in the 
control arm, and optionally a column defining subgroups for a subgroup analysis. 
If data are entered manually, then the meta-analysis and the forest plot on
the right are updated automatically with each complete row entered. If a row 
is incomplete or the data make no sense (for example, more events than 
participants randomized) no forest plot is drawn. Click "Add rows" to insert
more rows in the table widget, or "Clear empty rows" to clear all rows that 
are empty, except the last one (there is always a blank row at the bottom 
in order to fill in more data). 

Click the "Save to Excel" button to save the dataset as an Excel spreadsheet. 
Only the dataset is saved; in contrast, the "Export meta-analysis" button 
saves everything (dataset, analysis options, plot options, and the entire 
meta-analysis object) in an RDS file that can be imported in miniMeta, or 
further processed in R.

By clicking the "Show analysis options" checkbox, a suite of options appear
that determine how the meta-analysis is conducted (see the [documentation of 
the metabin() function](https://www.rdocumentation.org/packages/meta/versions/4.9-7/topics/metabin) in package [meta](https://CRAN.R-project.org/package=meta) for full details).
A fixed-effects or random-effects model can be specified, or both. 
There is a choice of different effect measures (Relative Risk, Odds Ratio, 
Risk Difference and Arcsine Difference), methods for pooling studies 
(Mantel-Hanszel, Inverse Variance and Peto) and heterogeneity estimators 
for the random-effects model (the default being the standard 
DerSimonian-Laird estimator). The Hartung and Knapp correction can be selected,
as well as the type of continuity correction for study arms with zero events, 
with a "treatment arm continuity correction" (TACC) being the default. 
More options may be added in the future.
With every change in these analysis options, the meta-analysis is 
automatically updated and a new forest plot is redrawn.


### Output section (right side)

The right section of the miniMeta window is where all the output appears. 
There are several tabs: "Forest plot", "Plot options", "GRADE output" 
and "Help". 

On the "Forest plot" tab, the forest plot appears automatically every time 
the input data or the analysis options are modified. 
Click the "Download plot" button to download the forest plot as an image file. 

On the "Plot options" tab, every aspect of the forest plot can be customized 
as desired. The options are conceptually organized in four sections: 
"Download options", "Content options", "Formatting options" and "Advanced 
options". Click on the corresponding checkbox for the options to appear.
"Download options" determine what kind of image file is downloaded
when clicking the "Download plot" button (pdf, ps, tiff or png), 
its dimensions (in inches), its resolution in dots per inch (dpi - only for 
bitmap-type images, i.e. tiff or png) and factors such as line width, spacing 
(between study rows) and pointsize (affecting the size of text).
"Content options" control what is shown in the forest plot and what is not, 
e.g. absolute numbers per study arm, weights, heterogeneity statistics, etc.
"Formatting options" affect various aesthetical characteristics of the 
forest plot or the funnel plot (see below). 

Finally, "Advanced options" provides a powerful interface (for advanced 
users) to control any aspect of the forest plot, by specifying a 
comma-separated list of arguments that are passed to the
[`forest.meta()`](https://www.rdocumentation.org/packages/meta/versions/4.9-7/topics/forest.meta) 
method. Any argument listed in the 
[function documentation](https://www.rdocumentation.org/packages/meta/versions/4.9-7/topics/forest.meta) 
can be specified in the input box, and these _override_ any other options 
set in miniMeta. For example, if `col.diamond="red"` is entered in the input
box, the diamond illustrating the pooled effect measure is painted red. 
This interface allows customizing the forest plot in ways that go beyond 
the options provided in miniMeta, but still without resorting to the R console.

The "GRADE output" tab provides results for a "Grading of 
Recommendations Assessment, Development and Evaluation" (GRADE) 
Summary of Findings (SoF) table. When filling in such a table,
each row corresponds to one meta-analysis of RCTs for a single outcome, 
and in the five rightmost columns one puts the Event Rate for the control
and intervention groups, the pooled relative effect (usually a Risk Ratio),
the absolute risk in the control group (per 1000 patients) and the absolute
Risk Difference in the intervention group. 
miniMeta provides these information automatically in the "GRADE output" tab.

The "Funnel plot" tab provides a funnel plot to assess publication bias or 
small study effects. The results for the Begg & Mamzudar and Egger tests 
are provided below the funnel plot, as well as a L'Abbe plot.

The last tab, "Help", includes some brief usage help to get you started 
with miniMeta.


## Observational studies module

The observational studies module roughly follows the interface of the 
RCT module, and most of the descriptions there apply here as well, 
but with some adaptations for the meta-analysis of observational studies.
Again the input data and analysis options are entered on the left side, 
with all output on the right side. "Import meta-analysis"/"Export 
meta-analysis" buttons can be found at the top. 

### Input section (left side)

Input (study) data can again be entered via an Excel file or on the 
table widget. However, the columns are different. The first column again 
contains the study labels, but the next columns contain the study effect 
point estimate, the lower/upper limits of the 95% Confidence Interval (95% CI), 
and the standard error (SE) of the effect. A final grouping column is 
available, if subgroup analyses are desired.

The table widget also operates in a slightly more complex way than 
in the RCT module. Studies normally report the point estimate of the 
effect and the associated 95% CI; however, the calculations 
in a meta-analysis of observational studies involve the point estimate and 
its _standard error_, which can be _derived_ from the confidence interval. 
As a result, while the effect measure and its 95% CI are entered in the 
table widget, the SE is automatically calculated and filled in the respective
table cell. SE always remains in sync with the provided 95% CI, unless the 
"Auto-calculate SE" checkbox is unchecked; then a SE can be entered manually 
(but, if blank, will still be automatically calculated based on the 95% CI). 
Potential problems with the data entered, such as SE that do not match the 
95% CI or a CI that are not symmetrical, are indicated with a message that 
appears just below the table widget.

Below the table widget is a drop-down menu there the type of effect measure 
is selected (Relative Risk, Odds Ratio, Risk Difference or Arcsine difference). 
This determines what kind of measure the "effect" and the "SE" column represent.
If the effect measure is the Relative Risk or Odds Ratio, then the SE column 
represents the standard error of the *log* Relative Risk or *log* Odds Ratio.

Click "Show analysis options" to adjust further options for the meta-analysis 
(see the [documentation of 
the metagen() function](https://www.rdocumentation.org/packages/meta/versions/4.9-7/topics/metagen) in package [meta](https://CRAN.R-project.org/package=meta) for full details).
A fixed-effects or random-effects model can be specified, or both. 
Different heterogeneity estimators can be chosen for the random-effects model 
(the default being the standard DerSimonian-Laird estimator), and the 
Hartung and Knapp correction can be selected as well.
With every change in these analysis options, the meta-analysis is 
automatically updated and a new forest plot is redrawn.

### Output section (right side)

The output section of the Observational studies module, on the right side of
the miniMeta window, is very similar to that of the RCT module. 

On the "Forest plot" tab, the forest plot appears automatically every time 
the input data or the analysis options are modified. 
Click the "Download plot" button to download the forest plot as an image file. 
Note that the 95% CIs on the forest plot *may not exactly correspond* to those 
on the input table widget. This is because what is analysed is the effect and 
its standard error, and the 95% CIs are recalculated based on that. 

On the "Plot options" tab, every aspect of the forest plot can be customized 
as desired. The four sections ("Download options", "Content options", 
"Formatting options" and "Advanced options") are the same as in the RCT module.
Some items in the "Content options" section that may not apply to observational 
studies have been omittted.

The "GRADE" output tab is again the same as in the RCT module, with one 
difference: an input field is provided to specify a baseline risk. 
This is because, in contrast to a meta-analysis of RCTs, the baseline risk 
cannot be estimated from the study data and must be plugged in separately. 
The "Funnel plot" section does not include a L'Abbe plot, as these are not 
applicable to observational studies.


## Tools 

At the "Tools" tab, a suite of small additional tools can be found that are 
useful when performing a meta-analysis. 
Currently this contains just one tool that implements the Bucher method for 
adjusted indirect comparisons.

### Bucher method tool

This tools implement the Bucher method for adjusted indirect comparisons. 
It utilizes two _pooled_ effect measures, comparing two outcomes against a 
common third outcome (outcome A vs outcome C, and outcome B vs outcome C).
Then a comparison of A vs B can be performed by taking the _difference_ 
of the two effect point estimates, and the _sum_ of their variances 
(squared standard errors). Just fill in the point estimate and 95% CI 
lower/upper limits for the two pooled effect estimates (A vs C, and B vs C), 
and the results will appear on the right side of the miniMeta window.


# Other miniMeta features

## Import/export functionality

In both the RCT module and Observational studies module, there are buttons 
for importing from / exporting to an RDS file. An RDS file holds a single 
serialized R object, and can be read into R using the console. The exported 
objects are of class 'miniMeta', and are essentially a list with four elements:
'data', 'meta', 'analysisOptions' and 'plotOptions'. Element 'data' contains 
the study data, as an R data.frame. Element 'meta' contains the meta-analysis 
(an object of class 'meta'). The other two elements are named lists containing 
all analysis and plot options. 

Objects of class 'miniMeta' can be further processed in R, or (when stored in an 
RDS file) can be imported back into miniMeta. They can also be converted into 
an R source code format using the `as.source()` function of the miniMeta 
package, and can be saved as an R script with something like 
`writeLines(as.source(m), "analysis.R")` (m is the 'miniMeta' class object).
Such a script contains (a) the data, (b) the command to run the meta-analysis,
and (c) the command to make the forest plot. It allows you to reproduce the 
meta-analysis, or extend it further into more complex analyses.


